A Framework for Equivalence of Cast Calculi and a new Cast Representation

### 2020-10-02 paper.tex
more comments

The semantics of these cast representations are defined in many
different papers with separate reduction rules, which makes it
difficult to isolate the key differences between them.  In this paper
we present the first set of reduction rules for a cast calculus that
is parameterized over the cast representation
(Section~\ref{sec:framework}). The semantics takes the form of a CEK
abstract machine and it formulates the interface between the machine
and the different cast representations as an abstract data type,
namely the \emph{Cast ADT}.

@@

The syntax and static semantics are the same for the \LD\ and
\LUD\ cast calculi. They are reviewed in \figref{fig:statics}.
As usual, the important features are the cast expressions,
$\ecast{e}{\ccast{S}{l}{T}}$, which are responsible for runtime type
checking. Blame labels can be positive (blaming the expression inside the cast)
or negative (blaming the context outside the cast), as in \cite{Wadler:2009qv,Siek:2015ab}.
The syntax and static semantics is the same as that of 
\citep{siek2009exploring} except for a few minor exceptions:

\begin{itemize}
\item We add product type.
\item We consider only one base type, Boolean, for simplicity.
\item We separate types into those with a type constructor at the top,
  the \emph{pre-types}, versus the dynamic type \Tdyn.
\end{itemize}
As usual, for every cast $\ecast{e}{\ccast{S}{l}{T}}$, the source type $S$ and 
the target $T$ must be consistent.

@@

The dynamic semantics of a cast calculus is typically defined with a
reduction semantics. Here we use a CEK machine
\citep{felleisen1986control} instead. One reason is that environments are 
easier 
to deal with than substitutions in formal proofs. Also, a CEK machine is more
convenient to use for the space-efficient semantics in
\chapref{sec:framework} (recall that a major point of space-efficiency 
is about compressing continuations and that continuations are part of the CEK 
machine states). So using a CEK machine for the cast 
calculi in this section makes it easier to prove these machines are equivalent.
Of course, one should also prove that the abstract machine
presented here is equivalent to the standard reduction semantics for
cast calculi, but we have not yet done so.

@@

Fig.~\ref{fig:R-machine-dynamics} defines the transition relation of the CEK
machine (\RMachine{B}) and Fig.~\ref{fig:R-machine-syntax} gives a grammar for machine 
states $s$, including a definition of values $v$ and value typing.
%
This machine is parameterized over blame strategy ($B$), which ranges over
Lazy D (Fig.~\ref{fig:blame-strategy-LazyD}) and Lazy UD 
(Fig.~\ref{fig:blame-strategy-LazyD}).
%
The transitions involving casts are highlighted in red and described in
more detail below. The other transitions are standard for a CEK
machine for an extended simply typed lambda calculus.
%
Recall that CEK machines involve two kinds of transitions: those
that dive further into an expression (looking for a redex) and push an
entry onto the continuation and those that return a value to the
current continuation and possibly perform a computation.
That is, the machine state is either in an evaluating
$\sexpr{e}{E}{K}$ or returning $\scont{v}{K}$ configuration, 
respectively. Additionally, there are two halting configurations: \shalt{v} 
halts with a value, while \error{l} halts with a blame label. 
Observables include all value constructors and blame labels, as in 
\citep{siek2012interpretations}. 
The function converting values to observables ($observe(v) = o$) is defined in 
the obvious way.

>>>>

The semantics of these cast representations are defined in many
different papers with separate reduction rules, which makes it
difficult to isolate the key differences between them.  In this paper
we present the first set of reduction rules for a cast calculus that
is parameterized over the cast representation
(Section~\ref{sec:framework}).
\todo[inline]{JS: First? What about \url{https://arxiv.org/abs/2001.11560}?}
The semantics takes the form of a CEK
abstract machine and it formulates the interface between the machine
and the different cast representations as an abstract data type,
namely the \emph{Cast ADT}.

@@

The syntax and static semantics are the same for the \LD\ and
\LUD\ cast calculi. They are reviewed in \figref{fig:statics}.
As usual, the important features are the cast expressions,
$\ecast{e}{\ccast{S}{l}{T}}$, which are responsible for runtime type
checking. Blame labels can be positive (blaming the expression inside the cast)
or negative (blaming the context outside the cast), as in \cite{wadler2009well}.
The syntax and static semantics is the same as that of 
\citep{siek2009exploring} except for a few minor exceptions:

\begin{itemize}
\item We consider only one base type, Boolean, for simplicity.
\item We add the product type.
\item We separate types into those with a type constructor at the top,
  the \emph{pre-types}, versus the dynamic type \Tdyn.
\end{itemize}
As usual, in a cast $\ecast{e}{\ccast{S}{l}{T}}$, the source type $S$
and the target $T$ must be consistent.

@@

The dynamic semantics of a cast calculus is typically defined with a
reduction semantics. Here we use a CEK machine
\citep{felleisen1986control} instead. One reason is that environments
are easier to deal with than substitutions in formal proofs. Also, a
CEK machine is more convenient to use for the space-efficient
semantics in \chapref{sec:framework} because continuations need to be
compressed and continuations are an explicit part of the CEK machine
states.  So also using a CEK machine for the cast calculi in this
section makes it easier to prove these machines are equivalent.  Of
course, one should also prove that the abstract machine presented here
is equivalent to the standard reduction semantics for cast calculi,
but we have not yet done so.

\todo[inline]{JS: We might cite the proof of correctness for CEK in the
  PLT Redex book and conjecture that it would be straightforward to
  adapt.}

@@

Fig.~\ref{fig:R-machine-dynamics} defines the transition relation of the CEK
machine (\RMachine{B}) and Fig.~\ref{fig:R-machine-syntax} gives a grammar for machine 
states $s$, including a definition of values $v$ and value typing.
\todo[inline]{JS: explain the name choice of $\mathcal{R}$.}
%
This machine is parameterized over blame strategy $B$, which ranges over
Lazy D (Fig.~\ref{fig:blame-strategy-LazyD}) and Lazy UD 
(Fig.~\ref{fig:blame-strategy-LazyUD}).
\todo[inline]{JS: I don't think $B$ is restricted to just those two.
Can't it be any blame strategy that satisfies certain conditions?}
%
The transitions involving casts are highlighted in red and described in
more detail below. The other transitions are standard for a CEK
machine for an extended simply typed lambda calculus.
%
Recall that CEK machines involve two kinds of transitions: those
that dive further into an expression (looking for a redex) and push an
entry onto the continuation and those that return a value to the
current continuation and possibly perform a computation.
That is, the machine state is either in an evaluating
$\sexpr{e}{E}{K}$ or returning $\scont{v}{K}$ configuration, 
respectively. Additionally, there are two halting configurations: \shalt{v} 
halts with a value, while \error{l} halts with a blame label. 
Observables include all value constructors and blame labels, as in 
\citep{siek2012interpretations}. 
The function converting values to observables ($observe(v) = o$) is defined in 
the obvious way.
